The multi-tenancy model has created two new security issues~\cite{dillon2010cloud}. Sharing resources on the same physical machine pose a danger to the data of the tenants. This falls within the isolation challenge. Another issue is reputation fate-sharing, since one might be sharing resources with possible criminal users, creating the possibility to (for example) get blacklisted.

They also mention the charging model as a challenge. The costs of developing multi-tenancy can be very substantial for \ac{SaaS} providers, because they need to re-design or re-develop single-tenant software, introduce new features for customization and enhance the security.
There are also benefits in terms of costs, as discussed by Guo et al.~\cite{guo2007framework} and Bezemer and Zaidman~\cite{bezemer2010multi}, since the applications have a higher utilization of resources. This results in lower overall costs, which also implicates lower prices for tenants. Bezemer and Zaidman~\cite{bezemer2010multi} additionally state that maintenance will also be cheaper (and easier) when multi-tenancy is implemented properly.

Bezemer and Zaidman~\cite{bezemer2010multi} also mention performance as a challenge, since multiple tenants are using the same hardware resources. Other challenges are scalability (usage of resources can suddenly increase), zero-downtime (the user expects the system to be online when he needs it) and maintenance (added complexity due to multi-tenancy could make code harder to maintain).